\blue{Numerical Investigation of Multi-Fluid Flows through Dual Saturated-Unsaturated Porous Media - change the title to something close to the main topic of the manuscript ... something with viscous flow instability or viscous fingering)}

### 2017-03-12 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
section 1 & vfingering
section 1 updated

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
\medskip
Numerical investigation and modeling of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterization of the flow of fluids through porous media serves as the foundation of reservoir engineering studies. Oil and gas migration and transport of contaminants in groundwater are some of important physical problems that porous media flows play a crucial role. 
%Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations.
Underground coal gasification, for the mining industry is an important area of interest, but more recently, also due to the global environmental problems, most research focused on the CO$_{\text{2}}$ migration and trapping mechanisms in carbon capture utilization and storage (CCUS) operations~\citep{spycher_2003, chen_2006, self_2012, pruess_1990c, white_1981,jiang_2011}.

\medskip
Due to complex and/or heterogeneous geological formations, multiphase flows through porous media play an important role in understanding the afformentioned operations. Modeling of multiphase flow, is an enormous technical and computational challenge. Starting from important contributions on the field of multiphase flows from \citet{wooding_1976}, \citet{homsy_1987} and \citet{adler_1988}, in the research area of fluid mechanics, there are the instabilities, also know as flow instabilities as these, first described by \citet{saffman_1958}. The first distinction should be made between microscopic and macroscopic instabilities. Microscopic instabilities are used for phenomena occurring locally, for example, Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}. On the other hand the macroscopic instabilities involves two-phase flow system \citep{bottoni_1992}. In this paper, the test-cases taht will follow focus on macroscopic instabilities and phenomena. A local flow instability that is the focus of this work is due to viscous and stress forces and often named as viscous instabilities or \textit{viscous fingering}.

\medskip
In large scale field operations, fingering is a relatively common condition in reservoirs with water-injection wells. As water and oil interacts, the interface between those two fluids moves along the geological formation, creating an uneven, or fingered, profile. The result of fingering is an inefficient sweeping action that can bypass significant volumes of recoverable oil and, in severe cases, an early breakthrough of water into adjacent production wellbores. This instability is mainly governed by the ratio of mobility of the displacing fluid and the of the displaced fluid (mobility-ratio, MR). %In a water-oil system the MR is given by eq.\ref{eq:MR}.
However, other important factors also influence the severity of viscous fingering: heterogeneous permeability fields, gravitational forces, anisotropic dispersion, velocity dependence of dispersion, non-monotonic viscosity profiles, etc.

@@

\medskip 
The conservation of mass along with the Darcy's law are the governing equations that describe the flow dynamics in porous media. Several computational methods have been developed to accurately represent force balanceand mass conservation in Darcean flows.
Finite difference methods (FDM) have been extensively used in most industry-standard flow simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications (e.g., black oil model, passive tracer advection etc). 
%The FDM discretization is based upon the differential form of the PDE to be solved. 
%Analytic models can be used to analyze reservoir performance as long as the reservoir is not highly heterogeneous. However, FDM are often limited to relatively simple geometries (with low grid quality and orientation) representing idealised geological formations~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}.
The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geometries. A finite element method (FEM) discretization is based upon a piecewise representation of the solution in terms of specified basis functions. The computational domain is divided up into smaller domains (finite elements) and the solution in each element is constructed from the basis functions. Advantages of FEM are the better description of the geometry and complex shapes, better treatment of fluid flow and saccurate more accurate stress calculations using a smaller number of nodes, less memory space and shorter run times. %At this point let us remind that a comparison of computaional methods is difficult and depends always on the test-case.
%Even though the most popular ones for the fluid flow equations are the Finite Difference Methods (FDM) there are also the Finite Volume Methods (FVM) and the Finite Element Methods. The latter is the one that will be used in this paper and more precisely a member of the FEM family, the Control-Volume FEM method (CVFEM).

\medskip
Among FEM-based formulations for porous media, the control volume finite element methods, ~\citep[CVFEM,][]{fung_1992} has been largely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements.

@@

In the CVFEM pressure is interpolated using piecewise linear finite element basis functions, %$\underline{\bf v}$ and $\underline{\bf p}$) 
while saturation is represented in node centered CV basis functions respectively, and the saturation equation is solved explicitly after solving for pressure at each time.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
\medskip
Numerical investigation and modeling of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterization of the flow of fluids through porous media serves as the foundation of reservoir engineering studies. Oil and gas migration and transport of contaminants in groundwater are some of important physical problems in which porous media flows play a crucial role. 
%Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations.
Underground coal gasification, for the mining industry is an important area of interest, but more recently, also due to global environmental problems, most research has focused on the CO$_{\text{2}}$ migration and trapping mechanisms in carbon capture utilization and storage (CCUS) operations~\citep{spycher_2003, chen_2006, self_2012, pruess_1990c, white_1981,jiang_2011}.

\medskip
Due to complex and/or heterogeneous geological formations, multiphase flows through porous media play an important role in understanding the afformentioned operations. Modeling of multiphase flow is an enormous technical and computational challenge. Important contributions regarding the description of physics and mechanisms related with multiphase flows have been made by \citet{wooding_1976}, \citet{homsy_1987} and \citet{adler_1988}. Nevertheless we should refer to the area of flow instabilities  as these, first described by \citet{saffman_1958}. The first distinction should be made between microscopic and macroscopic instabilities. Microscopic instabilities describe phenomena occurring locally, for example, Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}. On the other hand the macroscopic instabilities involves two-phase flow system \citep{bottoni_1992}. This paper will focus on microscopic instabilities and phenomena for a two-phase system. A local flow instability that is the focus of this work is due to viscous and stress forces and often named as viscous instabilities or \textit{viscous fingering}.

\medskip
In large scale field O$\&$G operations, viscous or density fingering is a relatively common condition in reservoirs with water-injection wells. As water and oil interacts, the interface between those two fluids moves along the geological formation, creating an uneven, or fingered, profile (see Fig.\ref{fig:simple_case}. The result of fingering is an inefficient sweeping action that can bypass significant volumes of recoverable oil and, in severe cases, an early breakthrough of water into adjacent production wellbores. This instability is mainly governed by the ratio of mobility (MR) of the displacing fluid and the of the displaced fluid (mobility-ratio, MR). %In a water-oil system the MR is given by eq.\ref{eq:MR}.
However, other important factors that also influence the severity of viscous fingering are: heterogeneous permeability fields, gravitational forces, anisotropic dispersion, velocity dependence of dispersion, non-monotonic viscosity profiles, etc.

@@

\medskip 
Conservation of mass along with the Darcy's law are the governing equations that describe the flow dynamics in porous media. Several computational methods have been developed to accurately represent force balances in Darcean flows.
Finite difference methods (FDM) have been extensively used in most industry-standard flow simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications (e.g., black oil model, passive tracer advection etc). 
%The FDM discretization is based upon the differential form of the PDE to be solved. 
%Analytic models can be used to analyze reservoir performance as long as the reservoir is not highly heterogeneous. However, FDM are often limited to relatively simple geometries (with low grid quality and orientation) representing idealised geological formations~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}.
The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geometrical domains. FEM discretization is based upon a piecewise representation of the solution in terms of specified basis functions. The computational domain is divided into smaller domains (elements) and the solution in each element is constructed from the basis functions. Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. %At this point let us remind that a comparison of computaional methods is difficult and depends always on the test-case.
%Even though the most popular ones for the fluid flow equations are the Finite Difference Methods (FDM) there are also the Finite Volume Methods (FVM) and the Finite Element Methods. The latter is the one that will be used in this paper and more precisely a member of the FEM family, the Control-Volume FEM method (CVFEM).

\medskip
Among FEM-based formulations for porous media, the control volume finite element methods, ~\citep[CVFEM,][]{fung_1992} has been largely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In the CVFEM pressure is interpolated using piecewise linear finite element basis functions, %$\underline{\bf v}$ and $\underline{\bf p}$) 
while saturation is represented in node centered CV basis functions respectively, and the saturation equations are solved explicitly after solving for pressure at each non-linear iteration.

@@
